### 2023-03-30 reachable-resilience.tex
details

\section{Resilience from a single-state for WSTS}

Resilience is a strong property since it implies that from every element of $\Bad$ there must exist a path to $\Safe$. But, when one considers a system with an initial state $s_0$, it could be sufficient to ask that only from $post^*(s_0) \cap \Bad$, there must exist a path to $\Safe$. However, this condition seems more difficult since one also must decide whether a state $s \in \Bad$ is reachable from $s_0$.
%
%\problemx{(I,J)-$k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis), a set $J$.}
%{$\forall s' \in J ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}
%
%\problemx{general $k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis).}
%{$\forall s'  ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}

Again, we first consider the case where the recovery time is bounded by a $k \in \N$.

\problemx{$k$-reachable-resilience problem for WSTS (kRRP)}
{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an upward-closed set $\Safe$ with a given basis, a decidable downward-closed set $\Bad$.}
{ $\Bad \cap \post^*(s) \longrightarrow^{\leq k} \Safe$ ?  \newline}
%

Analogously, we formulate bounded and general variations of the problem.

\problemx{bounded-reachable-resilience problem for WSTS (BRRP)}
{A state $s$ of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$, an upward-closed set $\Safe$ with a given basis, a decidable downward-closed set $\Bad$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is $k$-reachable-resilient. ?\newline}

\problemx{Rachable-resilience problem for WSTS (RRP)}
{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an upward-closed set $\Safe$ with a given basis, a decidable downward-closed set $\Bad$.}
{$ ~ \forall s' \in \Bad ~ (s \rightarrow^* s') \implies \exists s'' \in \Safe ~ s' \rightarrow^{*} s''$ ?\newline}

@@

\iffalse

\textcolor{blue}{
et encore: renforcer sur l'hypothèses sur I, J tous les deux clos par le haut+bas.
}

\textcolor{blue}{
compute the sets of $(I,J,k)$ st S is resilient.\\
max des I,J possibles \\
}

\fi

\begin{proposition}\label{reductions}
Coverability and $t$-liveness are reduceable to the reachable resilience problem.
%	home state and reachability
\end{proposition}

@@

\item $t$-liveness  can be expressed as the following formula
$ ~ \forall s' \in S ~ (s \rightarrow^* s') \implies \exists s'' \in U_t ~ s' \rightarrow^{*} s''$ 
where
$U_t=\uparrow \pred(t)$
is the upward closure of the preconditions to use transition $t$.  
The problem reduces itself to the reachable resilience problem
where $\Bad=S$ and $\Safe = \uparrow \pred(t)$.

@@

@@

\begin{theorem}\cite{DBLP:conf/gg/Ozkan22,DBLP:journals/corr/abs-2108-00889}
{\sc bounded-reachable-resilience} and {\sc $k$-reachable-resilience} are decidable for WSTS $S$ with strong upward compatibility and such that $\uparrow post^*$ is computable.
\end{theorem}

@@

\begin{corollary}\label{postcomputable}
{\sc reachable-resilience} is decidable for WSTS with strong upward compatibility and such that $\uparrow post^*$ is computable.
\end{corollary}

\begin{proof}
From Fact~\ref{stop condition}, there exists $k \in \N$ such that
$\pred^*(\Safe) =  \Safe^{k_m}$. We can compute this $k_m$ by iteratively computing
$\Safe^{k+1}$ from $\Safe^k$, checking $\Safe^{k+1}=\Safe^k$, 
returning $k_m$ if that is the case.
Then, because $k_m$-reachable-resilience is decidable, 
checking $\uparrow post^*(s) \cap \Bad \subseteq \Safe^k = \pred^*(\Safe)$ is,
and reachable-resilience is decidable.
\end{proof}

%\"Ozkan
\begin{proposition}[Proposition 1 in \cite{DBLP:conf/gg/Ozkan22}]
For finite-branching WSTS with strong upward compatibility, a basis of $\uparrow \post^*(s)$ is computable for every given state $s$ iff the downward-closed problem is decidable, i.e. given a state $s$ of a WSTS
% in the regarded class 
with strong upward compatibility 
and a decidable downward-closed set $D$, it can be decided whether $\exists s' \in D ~ s \to^* s'$. 
\end{proposition}

The characterization is used to show that Petri nets are $\post^*$-effective. It is well-known that Petri nets constitute WSTS with strong upward compatibility (Thm.~6.1 in \cite{DBLP:journals/tcs/FinkelS01}). 
Since VASS are WSTS with strong upward compatibility and since there is an algorithm that computes a finite basis of  $\uparrow \post^*(S)$, \cite{DBLP:conf/gg/Ozkan22} deduced that bounded reachable-resilience is decidable for VASS.
Hence reachable-resilience is decidable for Petri nets and VASS.
However, the hypothesis that $\uparrow \post^*$ is computable cannot be tested in the general WSTS framework.

@@

\begin{proof}
Let us show that Reset VASS don't enjoy the property that $\uparrow \post^*$ is computable.
Suppose that one are able to compute a finite basis of $\uparrow \post^*$ for Reset VASS. Then, one would be able to decide whether $0$ is reachable 
%in counter $i$ 
by examining if there is %some vector $v$ in the basis such that $v(i)=0$
$0$ in the basis%
. But reachability of $0$  %in a counter $i$ 
is undecidable for Reset VASS. 
% sont WSTS with strong compatibility. et l'accessibilité est undecidable.
%	$\uparrow post^*$ n'est pas calculable pour les LCS non plus car $\uparrow post^*= \uparrow \downarrow post^*$ n'est pas calculable ???
\end{proof}

>>>>

\section{State resilience}

Resilience is a strong property since it implies that from every element of $\Bad$ there must exist a path to $\Safe$. But, when one considers a system with an initial state $s_0$, it could be sufficient to ask that only from $\Bad \cap post^*(s_0)$, there must exist a path to $\Safe$. 
%
%			However, this condition seems more difficult since one also must decide whether a state $s \in \Bad$ is reachable from $s_0$.
%
%\problemx{(I,J)-$k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis), a set $J$.}
%{$\forall s' \in J ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}
%
%\problemx{general $k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis).}
%{$\forall s'  ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}

%		Again, we first consider the case where the recovery time is bounded by a $k \in \N$.

The three previous problems become:

\problemx{State-resilience problem (SRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$, $s \in S$ and two sets $\Safe, \Bad \subseteq S$.}
% 		an upward-closed set $\Safe$ with a given basis, a decidable downward-closed set $\Bad$.}
% 		{$ ~ \forall s' \in \Bad, ~ s \rightarrow^* s' \implies \exists s'' \in \Safe, ~ s' \rightarrow^{*} s''$ ?\newline}
{$\Bad \cap \post^*(s_0)  \rightarrow^{*} \Safe $ ? \newline}

\problemx{$k$-state-resilience problem (kSRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$, $s \in S$ and two sets $\Safe, \Bad \subseteq S$.}
{ $\Bad \cap \post^*(s) \longrightarrow^{\leq k} \Safe$ ?  \newline}
%

\problemx{bounded-state-resilience problem (BSRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$, $s \in S$ and two sets $\Safe, \Bad \subseteq S$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is $k$-state-resilient. ?\newline}

@@

%
\iffalse
%
\textcolor{blue}{
et encore: renforcer sur l'hypothèses sur I, J tous les deux clos par le haut+bas.
}
%
\textcolor{blue}{
compute the sets of $(I,J,k)$ st S is resilient.\\
max des I,J possibles \\
}
%
\fi

Since these problems are undecidable for general infinite-state transition systems, we restrict our study to WSTS.

\alain{mal placé}
\begin{proposition}\label{reductions}
In WSTS, coverability and $t$-liveness are reducible to the state resilience problem.
%	home state and reachability
\end{proposition}

@@

\item $t$-liveness  can be expressed as the following formula
$ ~ \forall s' \in S, ~ s \rightarrow^* s' \implies \exists s'' \in U_t, ~ s' \rightarrow^{*} s''$ 
where
$U_t=\uparrow \pred(t)$
is the upward closure of the preconditions to use transition $t$.  
The problem reduces itself to the reachable resilience problem
where $\Bad=S$ and $\Safe = \uparrow \pred(t)$.

@@

\problemx{Downward reachability problem}
{A state $s$, a downward-closed set $D$.}
{$s \to^* D$? \newline}

\begin{theorem}
The downward reachability problem is reducible to the state resilience problem.
\end{theorem}

\begin{proof}
Let $s \in S$, and a downward-closed set $D$.
Take $\Safe = \emptyset$ and $\Bad = D$.
The state resilience problem on $\Safe,\Bad$ asks
whether $\Bad \cap \post^*(s) \subseteq \emptyset$.
If the answer is negative then $s \to^* \Bad$,
else $s \not\to^* \Bad$.
\end{proof}

\alain{mal placé}

@@

Let us recall a result about state-resilience (called resilieence in \cite{DBLP:conf/gg/Ozkan22,DBLP:journals/corr/abs-2108-00889}).

\begin{theorem}\cite{DBLP:conf/gg/Ozkan22,DBLP:journals/corr/abs-2108-00889}
{\sc bounded-state-resilience} and {\sc $k$-state-resilience} are decidable for WSTS $S$ with strong upward compatibility and such that $\uparrow \post^*(s)$ is computable for $s \in S$.
\end{theorem}

@@

\begin{corollary}\label{postcomputable}
{\sc state-resilience} is decidable for WSTS with strong upward compatibility and such that $\uparrow \post^*(s)$ is computable for $s \in S$.
\end{corollary}

\alain{definir downward reachability problem.....
downward-closed problem given a state $s$ of a WSTS
% in the regarded class 
with strong upward compatibility 
and a decidable downward-closed set $D$, it can be decided whether $\exists s' \in D ~ s \to^* s'$. }

\begin{proof}
From Fact~\ref{stop condition}, there exists $k \in \N$ such that
$\pred^*(\Safe) =  \Safe^{k_m}$. We can compute this $k_m$ by iteratively computing \alain{mauvaises notations: $k, k_m, Safe^k,...$ n'a aucun sens...}
$\Safe^{k+1}$ from $\Safe^k$, checking $\Safe^{k+1}=\Safe^k$, 
returning $k_m$ if that is the case.
Then, because $k_m$-reachable-resilience is decidable, 
checking $\uparrow post^*(s) \cap \Bad \subseteq \Safe^k = \pred^*(\Safe)$ is,
and reachable-resilience is decidable.
\end{proof}

What are the WSTS for which $\uparrow \post^*(s)$ is computable for $s \in S$ ?
%\"Ozkan
\begin{proposition}[Proposition 1 in \cite{DBLP:conf/gg/Ozkan22}]
For finite-branching WSTS with strong upward compatibility, a basis of $\uparrow \post^*(s)$ is computable for every state $s$ iff the downward-reachability problem is decidable.
%i 			.e. given a state $s$ of a WSTS
% in the regarded class 
%with strong upward compatibility 
%and a decidable downward-closed set $D$, it can be decided whether $\exists s' \in D ~ s \to^* s'$. 
\end{proposition}
\alain{rappeler l'idée de la preuve}

The characterization is used to show that Petri nets are $\post^*$-effective. It is well-known that Petri nets are WSTS with strong upward compatibility (Thm.~6.1 in \cite{DBLP:journals/tcs/FinkelS01}). 
Since VASS are WSTS with strong upward compatibility and since there is an algorithm that computes a finite basis of  $\uparrow \post^*(S)$, \cite{DBLP:conf/gg/Ozkan22} deduced that bounded state-resilience is decidable for VASS.
Hence state-resilience is decidable for Petri nets and VASS.
However, the hypothesis that $\uparrow \post^*$ is computable cannot be tested in the general WSTS framework. Moreover, we may show:

@@

\begin{proof}
Let us show that Reset VASS, that are WSTSs with strong upward compatibility, don't enjoy the property that $\uparrow \post^*$ is computable.
Suppose that one are able to compute a finite basis of $\uparrow \post^*$ for Reset VASS. Then, one would be able to decide whether $0$ is reachable 
%in counter $i$ 
by examining if there is %some vector $v$ in the basis such that $v(i)=0$
$0$ in the basis%
. But reachability of $0$  %in a counter $i$ 
is undecidable for Reset VASS. 
% sont WSTS with strong compatibility. et l'accessibilité est undecidable.
%	$\uparrow post^*$ n'est pas calculable pour les LCS non plus car $\uparrow post^*= \uparrow \downarrow post^*$ n'est pas calculable ???
\end{proof}

### 2023-03-30 resilience-safe-down.tex
details

\begin{theorem}
The resilience problem is decidable for ideal-effective WSTS with 
$\Safe=\downarrow \Safe$
and $\Bad=\uparrow \Bad$
and
the additional hypothesis that
for all $D \subseteq S$ downward-closed, $\pred^*(D)$ is downward-closed.
% \alain{est-ce que  $\pred(D)$ is downward closed avec D downward closed serait suffisant ? si l'ordre est omega2 alors closbas+inclusion est wqo donc il faudrait juste que la suite croissante $pred^n(D)+le reste$ converge...}
\end{theorem}

\begin{proof}
By hypothesis $\pred^*(\Safe)$ is downward-closed, since $\Safe$ is downward-closed.
%  Let $\{J_1, J_2,...,J_n\}$ be the ideal decomposition of $\pred^*(\Safe)$ and $\{b_1,b_2,...,b_m\}$ be the (unique) minimal basis of $\Bad$.
The resilience problem can be reformulated as 
% $\uparrow \{b_1,b_2,...,b_m\} \subseteq J_1 \cup J_2 \cup \cdots \cup J_n$.
$\Bad \subseteq  \pred^*(\Safe)$.
Since $\mathscr{S}=(S,\rightarrow, \leq)$ is ideally effective, we can compute intersections of closed subsets.
Hence we can compute the intersection of
% $\uparrow \{b_1,b_2,...,b_m\}$
$\Bad$
and
$S \setminus \pred^*(\Safe)$,
which are both upward-closed.
Since
% $\uparrow \{b_1,b_2,...,b_m\} \subseteq J_1 \cup J_2 \cup \cdots \cup J_n$
$\Bad \subseteq \pred^*(\Safe)$
can be reformulated as
$\Bad \cap (S \setminus \pred^*(\Safe)) = \emptyset$
the resilience problem is decidable.
\end{proof}

\begin{corollary}
The resilience problem is decidable for ideal-effective downward-compatible WSTS with 
$\Safe=\downarrow \Safe$
and $\Bad=\uparrow \Bad$.
\end{corollary}

\begin{proof}
Downward compatibility of $\mathscr{S}=(S,\rightarrow, \leq)$ 
means that
for all $s_1, s_2, t_1 \in S$ with
$s_2 \leq s_1$ and $s_1 \to t_1$
there
exists
$t_2 \leq t_1$ such that $s_2 \to^* t_2$.

Let $D$ be a downward-closed subset of $S$
and let $x \in \downarrow \pred^*(D)$.
By design there exists
$y \in \pred^*(D)$ 
such that $x \leq y$
and $d \in D$ such that
$y = a_0 \to a_1 \to a_2 \to \cdots \to a_m \to a_{m+1} = d$.

By downward compatibility $a_0 \to a_1$
implies there exists $a'_1 \leq a_1$ with
$x \to^* a'_1$.
More generally $a_i \to a_{i+1}$ and
$a'_i\leq a_i$ implies the existence of $a'_{i+1} \leq a_{i+1}$ such that
$a'_i \to^* a'_{i+1}$,
and, by induction,
 $x \to^* a'_1 \to^* \cdots \to^* a'_{m} \to^* d'$
where
$d' \leq d$ belongs to $D$ by downward closure of $D$.
Hence $x \in \pred^*(D)$.
\end{proof}

% \begin{proposition}
In the case
of a ideal-effective WSTS 
where
the additional hypothesis that
for all $D \subseteq S$ downward-closed, $\pred^*(D)$ is downward-closed
is not met,
the above construction
can provide a proof
of non-resilience
i.e. when
$\Bad \cap (S \setminus \downarrow\pred^*(\Safe)) \neq \emptyset$
then
$\Bad \not\subseteq \downarrow\pred^*(\Safe)$
and hence
$\Bad \not\subseteq \pred^*(\Safe)$.
When $\Bad \cap (S \setminus \downarrow\pred^*(\Safe)) = \emptyset$
however
it is not enough to conclude.

@@

\begin{theorem}\label{down-down}
The resilience problem is undecidable for strong effective WSTS with
%	with more than one minimal element, 
$\Safe=\downarrow \Safe$
and $\Bad=\downarrow \Bad$.
\end{theorem}

\begin{proof}
Let us consider the case of a WSTS with at least two minimal elements $m_1$ and $m_2$.
The problem of whether $m_2$ is reachable from $m_1$ reduces itself to the resilience problem by considering $\Safe=\downarrow m_2 = \{ m_2\}$ and $\Bad=\downarrow m_1 = \{ m_1\}$. By undecidability of the reachability problem we conclude.  
\end{proof}

In the case of a WSTS with only one minimal element,  
then by downward closure the minimal element is present in both $\Safe$ and $\Bad$ which contradicts the assumption $\Safe \cap \Bad= \emptyset$.

>>>>

@@

\begin{theorem}\label{down-down}
The resilience problem is undecidable for effective WSTS with strong compatibility
%	with more than one minimal element, 
$\Safe=\downarrow \Safe$
and $\Bad=\downarrow \Bad$.
\end{theorem}

\begin{proof}
In the case of a WSTS with at least two minimal elements $m_1$ and $m_2$, the problem of whether $m_2$ is reachable from $m_1$ reduces itself to the resilience problem, hence its undecidability.  
\mathieu{dévellopper un peu la preuve, $m_2$ minimal de $\Safe$, $m_1$ minimal de $\Bad$...}
\end{proof}

In the case of a WSTS with only one minimal element,  
% (ex: Petri Nets and their «Zero») 
then by downward closure the minimal element is present in both $\Safe$ and $\Bad$ and hence resilience always hold.

\begin{theorem}\label{up-down}
\textcolor{red}{
The resilience problem is undecidable for WSTS with 
%	more than one minimal element, 
$\Safe=\downarrow \Safe$
and $\Bad=\uparrow \Bad$.
}
\end{theorem}

\begin{proof}
TBD
\end{proof}

\begin{theorem}
The resilience problem is decidable for WSTS with 
%	more than one minimal element, 
$\Safe=\downarrow \Safe$
and $\Bad=\uparrow \Bad$
and
\textcolor{red}{
additional hypothesis such that 
}
\alain{est-ce que  $\pred(D)$ is downward closed avec D downward closed serait suffisant ? si l'ordre est omega2 alors closbas+inclusion est wqo donc il faudrait juste que la suite croissante $pred^n(D)+le reste$ converge...}
\end{theorem}

\begin{proof}
Let $\{J_1, J_2,...,J_n\}$ be the ideal decomposition of $\Safe$ and $\{b_1,b_2,...,b_m\}$ be the (unique) minimal basis of $\Bad$.

Let's assume for now for simplicity's sake that $\Bad = \uparrow b$

As an hypothesis, we assume that $\pred^*(\Safe)$ is downward closed as well.
And let $\{J'_1, J'_2,...,J'_n\}$ be the ideal decomposition of $\pred^*(\Safe)$.

Resilience asks whether 
$ \uparrow b \subseteq J'_1 \cup J'_2 \cup ... \cup J'_n$.

Of note is that $ \downarrow b \subseteq J'_1 \cup J'_2 \cup ... \cup J'_n$
if and only if
there exists some $i$ such that
$\downarrow b \subseteq J'_i$.

\mathieu{Try and find a condition under which "$ \uparrow \{b\} \subseteq J'_1 \cup J'_2 \cup ... \cup J'_n$?" is decidable and/or find an undecidable problem that reduce to a variation on 
"$\uparrow \{b\} \subseteq J'_i$ ?"}
\end{proof}

### 2023-03-30 resilience-safe-up.tex
swapping subsections around
details

\begin{theorem}\label{up-up}
The resilience problem is decidable for effective WSTS with effective pred-basis, $\Safe=\uparrow \Safe$
and $\Bad= U \setminus \Safe$.

@@

>>>>

@@

\begin{theorem}\label{up-up}
The resilience problem is decidable for effective WSTS with effective pred-basis, $\Safe=\uparrow \Safe$
and $\Bad= U \setminus \Safe$ with $U=\uparrow U$.

### 2023-03-30 resilience.tex
details

\problemx{resilience problem (RP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two finite encodings of sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{*} \Safe$ ?\newline}
%
%\alain{il faudrait ne pas répéter 3 fois les mêmes imputs pour les 3 pbs: énoncer les 3 uniformes pbs d'un coup avec une fois l'input puis les 3 pbs pour un état s donné sans répéter non plus 3 fois les mêmes inputs}

@@

\problemx{$k$-resilience problem (kRP)}
{A transition system $\mathscr{S}=(S,\rightarrow), k \in \mathbb{N}$ and two finite encodings of sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{\leq k} \Safe$ ?\newline}

\problemx{bounded resilience problem (BRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two finite encodings of sets $\Safe, \Bad \subseteq S$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is %uniformely
 $k$-resilient ?\newline}

>>>>

\problemx{resilience problem (RP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{*} \Safe$ ?\newline}
%
%\alain{il faudrait ne pas répéter 3 fois les mêmes imputs pour les 3 pbs: énoncer les 3 uniformes pbs d'un coup avec une fois l'input puis les 3 pbs pour un état s donné sans répéter non plus 3 fois les mêmes inputs}

@@

\problemx{$k$-resilience problem (kRP)}
{A transition system $\mathscr{S}=(S,\rightarrow), k \in \mathbb{N}$ and two sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{\leq k} \Safe$ ?\newline}

\problemx{bounded resilience problem (BRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $\Safe, \Bad \subseteq S$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is %uniformely
 $k$-resilient ?\newline}
